---
title: "Recalibrating gene expression fold changes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recalibrating gene expression fold changes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Workflow

```{r setup}
library(recalibrate)
```

The basic recalibration workflow is very simple. A data.frame containing
differential expression per gene (measured as log fold change) is adjusted
using the variance in gene expression metric (V^G).

## Input data

The are two requirements for the input data.frame: The row names have to be
Ensembl gene ids, and one of the columns must contain log fold changes.

As an example, here we are using the first three rows of the Alasoo et al. 2018
dataset of macrophages stimulated with IFNg, which can be downloaded from
[zenodo](https://zenodo.org/records/839011)

```{r}
df <- data.frame(
  log2FoldChange = c(0.304144565986564, 2.0260124809332, 2.50289598633734),
  row.names = c("ENSG00000152213", "ENSG00000154642", "ENSG00000161929")
)
```

## Recalibration

The actual recalibration is then a single function call that returns the
recalibrated fold changes in column `recalibratedFC` of the same data.frame.
Note that the parameter `FC_col_name` needs to be adjusted if nominal fold
changes are not stored in column `log2FoldChange` (DESeq2 default).

```{r}
recalibrateFoldChange(df)
```

# Follow-up analysis example

As an example, here we use the Alasoo et al. 2018 dataset of macrophages
stimulated with IFNg [downloaded from zenodo](https://zenodo.org/records/839011).

```{r, include = FALSE}
datafile <- "../../data/naive_vs_IFNg_DESeq2_fold_change.txt.gz"
```
```{r}
library(ggplot2)
df <- read.table(datafile, header = TRUE, row.names = "gene_id")
df <- recalibrateFoldChange(df, remove_NA = TRUE)
ggplot(data = df, aes(y = recalibratedFC, x = log2FoldChange)) +
  theme_bw() +
  geom_point(alpha = 0.5, size = 1.2) +
  ylab(bquote("recalibrated fold change (log FC /" ~ sqrt(V^G) ~ ")")) +
  xlab("nominal fold change (log FC)")
```

In our analyses, we generally find a high degree of correlation between nominal 
and recalibrated fold changes.

```{r}
cor(abs(df[, c("log2FoldChange", "recalibratedFC")]),
  method = "spearman",
  use = "complete.obs"
)
```

## GO term enrichement comparison

TODO

# Acknowledgements

We would like to thank Sarah Kim-Hellmuth, Xiaoting Li, Ryan Collins,
Sanna Gudmundsson, Paul Hoffman, Mariia Mineava, Kaushik Ram Ganapathy,
and the current and former members of the Lappalainen Lab for helpful
discussions and code sharing.

# Funding

This work was supported by funding from the European Research Council (ERC)
under the European Union’s Horizon 2020 research and innovation programme
(Grant agreement no. 101043238) and the National Human Genome Research Institute
of the NIH (Grant no. R01GM140287).
Part of the computations were enabled by resources provided by the Swedish
National Infrastructure for Computing (SNIC) at UPPMAX partially funded by the
Swedish Research Council through grant agreement no. 2018-05973.

# Citation

> Rentzsch P, Kollotzek A, Mohammadi P, Lappalainen T (2024)
> Recalibrating differential gene expression by genetic dosage variance prioritizes functionally relevant genes.
> *bioRxiv*,
> [10.1101/2024.04.10.588830](https://doi.org/10.1101/2024.04.10.588830)
